Predicting complex diseases: performance, robustness and ..utility..(impact?)

### 2017-10-30 3-simus.Rmd
save

```{r}
PRS <- function(G, infos.chr, infos.pos, 
                pheno.all, covar.all, 
                ind.train, ind.test) {

@@

```{r}
params.grid2 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:100,
  stringsAsFactors = FALSE
) 
```

>>>>

```{r}
paste(sprintf("%.1e", 10^(-c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100))))), collapse = ", ")
PRS <- function(G, infos.chr, infos.pos, 
                pheno.all, covar.all, 
                ind.train, ind.test) {

@@

```{r}
params.grid2 <- expand.grid(
  par.causal = list(c(30, "all"), c(300, "all"), c(3000, "all"), c(30, "HLA")), 
  par.dist   = c("gaussian", "laplace"), 
  par.h2     = c(0.5, 0.8), 
  par.model  = c("simple", "fancy"),
  num.simu   = 1:50,
  stringsAsFactors = FALSE
) 
```

### 2017-10-30 paper/draft.tex
save

The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), called summary statistics.
For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (P) so that there remains only loci that are weakly correlated with each other. 
Thresholding (T) consists in removing SNPs that are under a certain level of significance (p-value threshold to be determined). 
Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb. [thresholds in results?] All these steps are implemented in R packages bigstatsr and bigsnpr [DIRE CA?? REF MOI]. We call this method ``PRS'' in the results.

@@

To fit this penalized logistic regression, we use a very efficient algorithms based on \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr. This type of algorithms build predictions for many values of $\lambda$ (typically 100). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this method separates the training set in $K$ folds (e.g. 10). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g. AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are then combined into one vector (e.g. using the geometric median). Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a linear combination of allele counts. We call this method ``logit-simple'' in the results.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard method, described previously. We call this method ``logit-triple'' in the results.

@@

First, we ran only 5 simulations for each combination of parameters (see section \ref{sec:simus}), excepted for the heritability that we fixed at $0.8$.
The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores. Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and have non-linear effects (Figure \ref{fig:ttrees}).
So, we decided to discard this method for the rest of the simulations.

@@

For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values from $10^{-0.1}$ to $10^{-100}$ equally spaced on the log-log-scale (Table \ref{tab:thr}). Note that, normally, the optimal threshold would have to be learned on the training set only. We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.

@@

First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}). So, we report only predictive performance in terms of AUC in following results.

As for the comparison of our methods with the P+T method, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario. Our penalized regression method consistently provides better predictive performance than the P+T method. Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores). Moreover, it results in sparse models (...).
The P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others and then the resulting summary statistics are used to construct a PRS.

@@

In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance as reported in previous studies [REF GAD + AUTRE?] \cite[]{abraham2013performance}, better than with the P+T method.
We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions. 
Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only.

>>>>

The P+T (Pruning + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), called summary statistics.
For the P+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (P) so that there remains only loci that are weakly correlated with each other. 
Thresholding (T) consists in removing SNPs that are under a certain level of significance (p-value threshold to be determined). 
Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ with regions of 500kb.

For the P+T method, we report three different scores of prediction: one including all the SNPs (after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximize the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}). Note that, normally, the optimal threshold would have to be learned on the training set only. We consider this reported maximum AUC as an upper bound of the AUC for the P+T method.
We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.

@@

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient}, also implemented in our package bigstatsr \cite[]{prive2017efficient}. This type of algorithms build predictions for many values of $\lambda$ (typically 100). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this method separates the training set in $K$ folds (e.g. 10). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes a criterion (e.g. AUC) is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are combined into one vector (e.g. using the geometric median). Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a linear combination of allele counts. We call this method ``logit-simple'' in the results.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct another dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times more variables than the initial one, on which we can apply the standard logistic regression with the CMSA procedure, described previously. We call this method ``logit-triple'' in the results.

@@

First, we ran only 5 simulations for each combination of parameters), excepted for the heritability that we fixed at $0.8$ (see section \ref{sec:simus}.
The goal was to quickly discard the T-Trees method as being competitive with the other methods. 
For example, compared to our penalized logistic regression, T-Trees perform worse for both predictive measures, the AUC and the percentage of cases in highest scores (Figure \ref{fig:ttrees}). Moreover, T-Trees takes longer to run and makes more complex predictive models because it uses more SNPs in the models and have non-linear effects (Figure \ref{fig:ttrees}).
So, we decided to discard this method for the rest of the simulations.

@@

@@

First, we show that AUC is highly correlated ($r^2 > 95\%$) with the percentage of cases in either the 10\% or 20\% largest PRSs (Figure \ref{fig:AUC-corr}). Therefore, AUC can be a good measure at determining whether a score can identify high-risk individuals or not, so we report only predictive performance in terms of AUC in following results.

Using the penalized logistic regressions, we report AUCs greater than 95\%, which is not obtained by the upper bound of the P+T method in any scenario. Our penalized regression methods consistently provides better predictive performance than the P+T method for all disease architectures tested. Our method is parallelized and requires only less than 10 minutes to run (with 6 physical cores). Moreover, it results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs.
The P+T runs in only a few seconds because we don't consider the GWAS computation as part of the execution time of the method because this is generally executed by others and then the resulting summary statistics are used to construct a PRS.

@@

In this comparative study, we proposed other methods as a replacement to the widely used P+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really good performance, as reported in previous studies \cite[]{abraham2013performance}, better than with the P+T method.
We are well aware that the P+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions. 
Yet, in the case of prediction of disease to be used for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only.
